---
title: "Introduction to mitolina: MITOchondrial LINeage Analysis"
output: rmarkdown::html_vignette
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Mikkel Meyer Andersen
bibliography: refs.bib
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, cache = FALSE)
```

## Introduction

First, the library is loaded:
```{r, message=FALSE}
library(mitolina)
```

For reproducibility, the seed for the (pseudo) random number generator is set:
```{r}
set.seed(1)
```

In this vignette, I will use functionality from the `tidyverse` (the pipe operator from the `magrittr` package and `dplyr` functionality). 
More information about these are available at <https://www.tidyverse.org/>, <https://magrittr.tidyverse.org/> and <http://dplyr.tidyverse.org/>.

```{r, message=FALSE}
library(tidyverse)
```

### Simulating genealogies

A genealogy is simulated by specifying population sizes for females and males.

```{r}
sim <- sample_mtdna_geneology_varying_size(
  population_sizes_females = rep(10, 10),
  population_sizes_males = c(0L, rep(10, 9)),
  progress = FALSE)
pop <- sim$population
pop
```

Note, that the first male generation must have size 0.

Please refer to the documentation of `sample_mtdna_geneology_varying_size` (`?sample_mtdna_geneology_varying_size`) for details in variability in reproductive success (VRS) [@AndersenPLOSGEN2017].

This only sample a genealogy. We also need to build pedigrees:

```{r}
peds <- build_pedigrees(pop, progress = FALSE)
peds
```


We can then plot one pedigree. We take a pedigree with a male individual in the final generation (`end_generation_male_individuals`):

```{r}
end_gen_peds <- lapply(sim$end_generation_male_individuals, get_pedigree_from_individual)
pedids <- sapply(end_gen_peds, get_pedigree_id)
id_tab <- sort(table(pedids))
ped_id_endgen <- as.integer(names(id_tab)[1]) # smallest
indv_index <- which(pedids == ped_id_endgen)[1]
ped <- get_pedigree_from_individual(sim$end_generation_male_individuals[[indv_index]])
pedigree_size(ped)
```

```{r}
plot(ped)
```

The squares are males and the circles are females.

### Adding mitogenomes

We are now ready to add haplotypes/mitogenomes to the individuals. Let us consider 
```{r}
sites <- 100
```
in the mitogenome for know.

First, we construct a function to create founder haplotypes. Here, they are just random:

```{r}
get_founder_mito <- function() {
  sample(c(FALSE, TRUE), sites, replace = TRUE)
}
x <- get_founder_mito()
head(x)
sum(x)
```

Let us assume the following mutation rates per site:

```{r}
mu <- rep(1e-2, sites)
```

Now, we can add mitogenomes:

```{r}
pedigrees_all_populate_haplotypes_custom_founders(
  pedigrees = peds,
  mutation_rates = mu,
  get_founder_haplotype = get_founder_mito, 
  progress = FALSE)
```

We can then try to plot the same pedigree again, this time with information about the individuals' mitogenomes (namely, the number of variants in each individuals mitogenome):

```{r}
plot(ped, num_vars = TRUE)
```

Plotting with the `igraph` plot quickly becomes difficult to customise. Therefore extensions to the `tidygraph`/`ggraph` packages is provided:

```{r, fig.height=8, fig.width=12}
library(tidygraph)
library(ggraph)

g <- as_tbl_graph(peds) %>% 
  activate(nodes) %>% 
  filter(ped_id == ped_id_endgen) %>% 
  mutate(num_vars = sapply(haplotype, sum))

p <- ggraph(g, layout = 'tree') +
  geom_edge_link() +
  geom_node_point(aes(color = sex, fill = num_vars, shape = sex), size = 12, stroke = 2) +
  geom_node_text(aes(label = paste0(name, "\n", num_vars)), size = 4, color = "white") +
  theme_graph(base_family = "") +
  scale_shape_manual(NULL, values = c("Female" = 21, "Male" = 22)) +
  scale_color_manual(NULL, values = c("Female" = "black", "Male" = "red"))
p
```

Here, the top number is the id and the number below is the number of variants in the individual's mitogenome.
Note, how it is possible to get both a (border) color and fill for `pch` values 21-25.

We can also extract all haplotypes in this pedigree:

```{r}
haps <- get_haplotypes_in_pedigree(ped)
length(haps)
mean(sapply(haps, sum))
```

Or from the last male generation by

```{r}
haps <- get_haplotypes_individuals(sim$end_generation_male_individuals)
length(haps)
```

In the simulation function `sample_mtdna_geneology_varying_size`, there is a parameter `extra_individuals_generations_return` to decide how many individuals are returned in the `male_individuals_generations` and `female_individuals_generations` elements. 
The default is `2`, meaning that individuals in 2 additional generations (on top of the final generation) are returned. 
So, with the default, the last three generations of males are available in the `male_individuals_generations` element. 

```{r}
haps <- get_haplotypes_individuals(sim$male_individuals_generations)
length(haps)
```


### mtDNA information included

There are several datasets with information about the mitogenome included.
There is a help file to each of the datasets which can be found by e.g. `?mtdna_partitions` for 
the `mtdna_partitions` dataset.

First, there is information about the positions from PhyloTree <http://www.phylotree.org/resources/rCRS_annotated.htm> about the revised Cambridge Reference Sequence (rCRS) [@rCRS]:

```{r}
data(mtdna_partitions)
mtdna_partitions %>% print(n = 2)
```

There is also mutation information from [@Soares2009]:

```{r}
data(mtdna_mut_SoaresK8)
mtdna_mut_SoaresK8
```

From [@Rieux2014]

```{r}
data(mtdna_mut_RieuxK4)
mtdna_mut_RieuxK4

data(mtdna_mut_RieuxK1)
mtdna_mut_RieuxK1
```

And from [@Oversti2017]:

```{r}
data(mtdna_mut_OverstiK4)
mtdna_mut_OverstiK4
```

All these are assembled info one dataframe:

```{r}
data(mtdna_mut_schemes)
mtdna_mut_schemes %>% print(n = 2)
mtdna_mut_schemes %>% count(MutationScheme)
```

This makes it easier to use the mutation rates from for example `mtdna_mut_RieuxK4`:

```{r}
d_mu <- mtdna_mut_schemes %>% 
  filter(MutationScheme == "Rieux 2014 (K = 4)", MutRateYearly > 0)
mu_mean_year <- d_mu %>% pull(MutRateYearly)
mu_mean_gen <- 25*mu_mean_year
```

Now, `mu_mean_gen` contains a mutation rate per generation (of 25 years) for each of the `r length(mu_mean_gen)` sites. The probability of one ore more mutations in a single meiosis is then

```{r}
1 - prod(1 - mu_mean_gen)
```

Note, that to include variability in the mutation rates, 
it is probably best to sample a rate per region (in `mtdna_mut_RieuxK4`), and then distribute 
that to all sites in the partition.


These rates are easily used to populate mitogenomes:

```{r}
get_founder_mito <- function() {
  sample(c(FALSE, TRUE), length(mu_mean_gen), replace = TRUE)
}

pedigrees_all_populate_haplotypes_custom_founders(
  pedigrees = peds,
  mutation_rates = mu_mean_gen,
  get_founder_haplotype = get_founder_mito, 
  progress = FALSE)
```

Number of variants in the last three male generations (not just in the one pedigree, but in entire population):

```{r}
haps <- get_haplotypes_individuals(sim$male_individuals_generations)
table(sapply(haps, sum))
```

Number of variants in the last male generation (not just in the one pedigree, but in entire population):

```{r}
haps <- get_haplotypes_individuals(sim$end_generation_male_individuals)
table(sapply(haps, sum))
```

And the same pedigree with new mitogenomes, now where number of variants is a factor to get a discrete fill colour scale:

```{r, fig.height=8, fig.width=12}
g <- as_tbl_graph(peds) %>% 
  activate(nodes) %>% 
  filter(ped_id == ped_id_endgen) %>% 
  mutate(num_vars = sapply(haplotype, sum))

p <- ggraph(g, layout = 'tree') +
  geom_edge_link() +
  geom_node_point(aes(color = sex, fill = factor(num_vars), shape = sex), size = 12, stroke = 2, show.legend = FALSE) +
  geom_node_text(aes(label = paste0(name, "\n", num_vars)), size = 4, color = "white") +
  theme_graph(base_family = "") +
  scale_shape_manual(NULL, values = c("Female" = 21, "Male" = 22)) +
  scale_color_manual(NULL, values = c("Female" = "black", "Male" = "red"))
p
```


## References
